Review Paper Draft

### 2020-08-15 main.tex
Section 1 edits

\anna{get to the point sooner}

@@

\pb{This paragraph seems like too much of a digression \ldots let's get to the point sooner} \anna{this could be more like a sentence} Genetic variants are often correlated with each other such that univariate testing may be prone to false positives and biased estimates. When a genetic variant of interest is highly correlated with others, the effects of these will be attributed to the variant in question. \anna{I feel like I'm saying 'variant' way too much... SNP? gene?} One method for analyzing data in the presence of correlated loci is LD pruning. LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods. The overarching approach of these methods is to sequentially thin SNPs, so that only SNPs correlated with each other below some desired threshold are used in modeling and testing procedures. This may result in causal SNPs being removed from analysis, so that only SNPs in LD with the causal loci, as opposed to the causal loci themselves, can be identified and their effects estimated.

% LD pruning thus limits analysis to SNPs which are representative of a particular region.

% In some scenarios pruning algorithms may not select any representative SNPs in a particular region of the genome \citep{prive2018efficient}. However, LD-pruning has been shown to improve performance of principal component analysis (PCA) methods \citep{abdellaoui2013population} and \anna{what} \anna{because why}.

An alternative solution to the problem of LD is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, classical approaches rely on collecting data with more observations than unknown parameters. This is typically infeasible in genome-wide studies where the number of features greatly exceeds the number of observations. Penalized regression methods make such problems tractable by constraining the solution space of coefficient estimates. The Lasso is a widely used penalized regression method \citep{tibshirani1996regression}. It relies on an assumption of sparsity, which allows coefficient estimation and variable selection to be accomplished simultaneously. It also provides considerable computational advantages that scale up well to very large data sets. These features make the Lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest.

@@

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods \pb{I really like this sentence \ldots can we work it into the abstract?}. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations \citep{Sillanpaeae2011, sul2018population}, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences \pb{do we review \emph{consequences} in section 3?} of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

>>>>

@@

% \anna{keep this for later use when talking about genetic background?} Genetic variants are often correlated with each other such that univariate testing may be prone to false positives and biased estimates. When a genetic variant of interest is highly correlated with others, the effects of these will be attributed to the variant in question. This may result in causal SNPs being removed from analysis, so that only SNPs in LD with the causal loci, as opposed to the causal loci themselves, can be identified and their effects estimated.

One method for analyzing data in the presence of correlated loci is LD pruning. LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods. The overarching approach of these methods is to sequentially thin SNPs, so that only SNPs correlated with each other below some desired threshold are used in modeling and testing procedures. An alternative solution to the problem of LD is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, classical approaches rely on collecting data with more observations than unknown parameters. This is typically infeasible in genome-wide studies where the number of features greatly exceeds the number of observations. Penalized regression methods make such problems tractable by constraining the solution space of coefficient estimates. The Lasso is a widely used penalized regression method \citep{tibshirani1996regression}. It relies on an assumption of sparsity, which allows coefficient estimation and variable selection to be accomplished simultaneously. It also provides considerable computational advantages that scale up well to very large data sets. These features make the Lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest.

@@

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations \citep{Sillanpaeae2011, sul2018population}, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences \pb{do we review \emph{consequences} in section 3?} \anna{tbd} of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section \ref{sec:results}, we illustrate the concepts described in section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.
